Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2023-12-04 manuscript/manuscript.qmd
starting on summary stuff

---
title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require an extraction process.  We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (r2 ranging from 0.78-0.96).  We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (r2 between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
#filters: 
#  - authors-block
---

@@

To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots to assess the fit and compare the different sensors. We derived the coefficient of determination from the regressions as a measure of fit and qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts. We do not compare slopes as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.

@@

All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll a. All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship. Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran. Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in-vivo concentrations of chlorophyll (Zamyadi et al . 2012). Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments. A study by Silva et al. found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al. 2016).

This study supports the use of extracted chlorophll and phycocyanin as an alternative to cell counts for the study waterbodies. Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments. Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts. There are a variety of instruments capable of measuring phycocyanin using in-vivo fluorometry across a spectrum of prices. We hope this study will encourage further research and consideration of using in-vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.

>>>>

---
title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require an extraction process.  We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (r2 ranging from 0.78-0.96).  We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (r2 between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
---

@@

To compare extracted concentrations and cell counts/biovolume to the various instruments we used simple linear regression and scatterplots. We derived the coefficient of determination from the regressions as a measure of fit and also qualitatively compare the device measurements to the extracted concentrations and phytoplankton counts with scatterplots. We do not compare slopes nor look for a slope of 1 as we have little expectation that the extracted concentrations would equal the values from the various devices given the difference in units and the different RFU returns across the devices.

@@

All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll a. All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship. Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran. Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (Zamyadi et al . 2012). Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments. A study by Silva et al. found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al. 2016).

This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies. Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments. Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts. There are a variety of instruments capable of measuring phycocyanin using in-vivo fluorometry across a spectrum of prices. We hope this study will encourage further research and consideration of using in-vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.
